---
title: "Static and Interactive Maps"
editor_options: 
  chunk_output_type: console
---

These are a few plots I made as a part of a homework assignment for my data science 2 class! The data I used is from the `poliscidata` package.

```{r}
#| message: false
#| warning: false
#| echo: false

# load needed libraries

library(tidyverse)
library(leaflet)
library(sf)
library(maps)
library(poliscidata)
library(viridis)
library(glue)
library(htmltools)
```

```{r}
#| message: false
#| warning: false
#| echo: false

# Get info to draw US states for geom_polygon 
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# Get info to draw US states for geom_sf and leaflet 
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# Load in state-wise data for filling our choropleth maps
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)

```

```{r}
#| echo: false
#| message: false

# Clean all data sets in preparation of merging them

## states_polygon
states_polygon <- states_polygon |>
  rename(state = region) |>                        # needs consistent col. name
  filter(!state %in% c("district of columbia")) |> # remove inconsistent region 
  mutate(state = str_replace_all(state, " ", ""))  # no more spaces!

## state_sf
states_sf <- states_sf |> 
  mutate(name = str_replace_all(name, " ", "")) |> # no more spaces!
  rename(state = name) |>                          # consistent name 
  filter(!state %in% c("Alaska", 
                       "Hawaii", 
                       "DistrictofColumbia",       # remove inconsistent regions
                       "PuertoRico")) |>
  mutate(state = str_to_lower(state))              # lowercase to match others

## polisci
polisci_data <- polisci_data |>                    
  mutate(state = str_squish(state)) |>             # no more spaces
  filter(!state %in% c("Alaska", "Hawaii")) |>     # remove inconsistent regions
  mutate(state = str_to_lower(state))              # lowercase to match others

```

```{r}
#| echo: false
#| message: false
# Merge polisci with states_polygon (for static)
gg_states <- polisci_data |>
  full_join(states_polygon)

# Check that merge worked for 48 contiguous states
#   unique(gg_states$state)

# Merge polisci with states_sf (for interactive)
int_states <- polisci_data |>
  full_join(states_sf)

# Check that merge worked for 48 contiguous states
#   unique(int_states$state)
int_states <- st_as_sf(int_states)

# fix capitalization and spacing of states for interactive plot
int_states <- int_states |> 
  mutate(state = str_replace(state, "new", "New "),
         state = str_replace(state, "north", "North "),
         state = str_replace(state, "south", "South "),
         state = str_replace(state, "rhode", "Rhode "),
         state = str_replace(state, "west", "West "),
         state = str_to_title(state)
         ) 
# class(int_states)
```

## Motor Vehicle Fatalities Across America

### Static Map

Here, you can see that Mississippi and Montana had the most car fatalities per 100,000 in 2007 as they are the lightest in color. The Northeast region of the U.S. seems to have the least amount of car fatalities per 100,000 as they are the darkest states.

```{r}
#| echo: false
gg_states |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = carfatal07), color = "black") + 
  coord_map() +
  scale_fill_viridis() +                              # better colors
  theme_void() +                                      # cleaner theme
                                                        # removes x and y scale
  labs(fill = "Motor vehicle fatalities per 100,000 pop.",
       caption = "Data from the poliscidata R package",
       title = "Motor vehicle fatalities per 100,000 people in 2007")
```

### Interactive Map

In the interactive version, you can see exactly how many average car fatalities occur in each state instead of inferring from the color and legend.

```{r}
#| echo: false

# set bins for fill colors
bins <- c(0, 5, 10, 15, 20, 25, 30, 35) 

# assigns fill colors to variable
pal <- colorBin("YlOrRd", domain = int_states$carfatal07, bins = bins)

# create labels
int_states <- int_states |> 
  mutate(labels = str_c(state, ": ", 
                        carfatal07, " motor vehicle fatalities / 100,000 pop"))
labels <- lapply(int_states$labels, HTML)

# leaflet plot
leaflet(int_states) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(carfatal07),
    weight = 2,                             # editing lines separating states...
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 1,
    highlightOptions = highlightOptions(    # options for hovering over states
      weight = 5,                             # editing hover lines
      color = "white",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),                   # brings hover line to front
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"), # font edits
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal, values = ~carfatal07, opacity = 0.7, title = NULL,
    position = "bottomright")
  
```

## Political Party Lean Across America

### Static Map

In this plot, you can generally see where states and regions of the U.S. lean politically. Here, you can infer what states are potential "swing" states in presidential elections. It is also interesting to see how the coasts lean more democratic compared to the midwest and south lean more republican.

```{r}
#| echo: false
gg_states |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = cook_index3), color = "black") +
  coord_map() +
  scale_fill_manual(values = c("red", "purple", "blue")) +
  theme_void() +
  labs(fill = "Cook Index",
       caption = "Data from the poliscidata R package",
       title = "Political party lean by state")
```

# Interactive Map

You can see a lot of the same information from the static plot in this interactive map.However, with the interactive version, you can zoom in/hover on the smaller states (RI, NJ, etc) to see exactly where they lean politically more easily than you can on the static map.

```{r}
#| echo: false

# fill colors
colors <- colorFactor(c("red", "purple", "blue"), domain = int_states$cook_index3)

# create labels
int_states <- int_states |>
  mutate(labels2 = str_c(str_to_title(state), " : ", cook_index3))
labels2 <- lapply(int_states$labels2, HTML)

# leaflet plot
leaflet(int_states) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~colors(cook_index3),
    weight = 2,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "lightgrey",
      dashArray = "",
      bringToFront = TRUE),
    label = labels2,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = colors, values = ~cook_index3, opacity = 0.7, title = NULL,
    position = "bottomright") 

```
