---
title: "Static and Interactive Maps"
editor_options: 
  chunk_output_type: console
---

These are a few plots I made as a part of a homework assignment for my data science 2 class! The data I used is from the `poliscidata` package.

```{r}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(leaflet)
library(sf)
library(maps)
library(poliscidata)
library(viridis)
library(glue)
library(htmltools)
```

```{r}
#| message: false
#| warning: false
#| echo: false

# Get info to draw US states for geom_polygon (connect the lat-long points)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# Get info to draw US states for geom_sf and leaflet (simple features object 
#   with multipolygon geometry column)
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# Load in state-wise data for filling our choropleth maps
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)

```

## Motor Vehicle Fatalities Across America

```{r}
#| echo: false
#| message: false
# states_polygon
states_polygon <- states_polygon |>
  rename(state = region) |>
  filter(!state %in% c("district of columbia")) |>
  mutate(state = str_replace_all(state, " ", ""))

# state_sf
states_sf <- states_sf |> 
  mutate(name = str_replace_all(name, " ", "")) |>
  rename(state = name)

states_sf <- states_sf |>
  filter(!state %in% c("Alaska", "Hawaii", "DistrictofColumbia", "PuertoRico")) |>
  mutate(state = str_to_lower(state))

# polisci
polisci_data <- polisci_data |> 
  mutate(state = str_squish(state)) |>
  filter(!state %in% c("Alaska", "Hawaii")) |>
  mutate(state = str_to_lower(state))

```

```{r}
#| echo: false
#| message: false
# Merge with states_polygon (static)
gg_states <- polisci_data |>
  full_join(states_polygon)

# Check that merge worked for 48 contiguous states
#   unique(gg_states$state)

# Merge with states_sf (static or interactive)
int_states <- polisci_data |>
  full_join(states_sf)

# Check that merge worked for 48 contiguous states
#   unique(int_states$state)
int_states <- st_as_sf(int_states)
# class(int_states)
```

### Static Map

Here, you can see that Mississippi and Montana had the most car fatalities per 100,000 in 2007 as they are the lightest in color. The Northeast region of the U.S. seems to have the least amount of car fatalities per 100,000 as they are the darkest states.

```{r}
#| echo: false
gg_states |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = carfatal07), color = "black") +
  coord_map() +
  scale_fill_viridis() +
  theme_void() +
  labs(fill = "Motor vehicle fatalities per 100,000 pop.",
       caption = "Data from the poliscidata R package",
       title = "Motor vehicle fatalities per 100,000 people in 2007")
```

### Interactive Map

In the interactive version, you can see exactly how many average car fatalities occur in each state instead of inferring from the color and legend.

```{r}
#| echo: false

bins <- c(0, 5, 10, 15, 20, 25, 30, 35)
pal <- colorBin("YlOrRd", domain = int_states$carfatal07, bins = bins)
int_states <- int_states |> 
  mutate(labels = str_c(str_to_title(state), ": ", carfatal07, " motor vehicle fatalities / 100,000 pop"))
labels <- lapply(int_states$labels, HTML)

leaflet(int_states) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(carfatal07),
    weight = 2,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 1,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "red",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal, values = ~carfatal07, opacity = 0.7, title = NULL,
    position = "bottomright")
  
```

## Political Party Lean Across America

### Static Map

In this plot, you can generally see where states and regions of the U.S. lean politically. Here, you can infer what states are potential "swing" states in presidential elections. It is also interesting to see how the coasts lean more democratic compared to the midwest and south lean more republican.

```{r}
#| echo: false
gg_states |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = cook_index3), color = "black") +
  coord_map() +
  scale_fill_manual(values = c("red", "purple", "blue")) +
  theme_void() +
  labs(fill = "Cook Index",
       caption = "Data from the poliscidata R package",
       title = "Political party lean by state")
```

# Interactive Map

You can see a lot of the same information from the static plot in this interactive map.However, with the interactive version, you can zoom in/hover on the smaller states (RI, NJ, etc) to see exactly where they lean politically more easily than you can on the static map.

```{r}
#| echo: false
colors <- colorFactor(c("red", "purple", "blue"), domain = int_states$cook_index3)

int_states <- int_states |>
  mutate(labels2 = str_c(str_to_title(state), " : ", cook_index3))
labels2 <- lapply(int_states$labels2, HTML)

leaflet(int_states) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~colors(cook_index3),
    weight = 2,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "black",
      bringToFront = TRUE),
    label = labels2,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = colors, values = ~cook_index3, opacity = 0.7, title = NULL,
    position = "bottomright") 

```
